Gradual Soundness: Lessons from Static Python

### 2022-01-19 pj-7.1/paper.tex
pj: comment from carl about fastlen

\newcommand{\numbenchmark}{2}
\newcommand{\numSPtest}{214}
\newcommand{\CPUchange}{3.7\%}

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Pyre~\cite{pyre}, Nom~\cite{mt-oopsla-2017,mt-oopsla-2021}, and PEP
484~\cite{pep484} are three notable sources.
In addition, \SP{} is strongly influenced by the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions and first-class classes
    are thus a low priority; instead, the dynamic type is available as a coarse
    approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of Pyre
    type annotations.
    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
    % of types that enable optimizations.
    % depth-first types.
    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated with constant-time casts.
Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.

@@

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
module imports, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's random testing tools~\cite{kf-sfp-2009} to check
type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
%% FILL how many programs exactly?
%% FILL what bugs did testing catch?
Finally, we employed the well-established method of testing end-to-end
conformance~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}
with a test suite.
In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,\footnote{For most of the \numSPtest{} tests, the translation is automatic.
  A few tests required hand-pruning to remove features that the model
  does not handle (\totalnum{52}).}
which suggests that the model conforms to actual \SP{}.

@@

\section{Fine-Grained Benchmark Data}

@@

\smallskip
{\centering\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}}

>>>>

\newcommand{\numbenchmark}{2}
\newcommand{\numSPtest}{214}
\newcommand{\numSPtotal}{802} %% SP commit 6d61575 from Jan 9
\newcommand{\numSPdiff}{588} %% (\numSPtotal - \numSPtest)
\newcommand{\CPUchange}{3.7\%}

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Pyre~\cite{pyre}, an optional type checker developed by Facebook,
greatly influenced the syntax and types of \SP{}.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021}, a research language with compelling
performance results, set the basic strategy for run-time checks.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions and first-class classes
    are thus a low priority; instead, the dynamic type is available as a coarse
    approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of Pyre
    type annotations.
    \SP{} is therefore compatible with Pyre syntax to reduce the
    adoption burden on its target users.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated with constant-time casts.
Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.

@@

%% TODO 2022-01-16 clarify why fast_len is interesting at all:
%% Carl: Two parts to the answer here. In the early days of Static Python we were
%% focused on matching Cython in some very hot code paths, and a couple of those
%% did take length of containers. So we implemented some things like FAST_LEN that
%% may not be broadly applicable but we needed to be as fast as possible in those
%% few Cython conversions. But since then we also implemented a general
%% optimization for iteration of loops where the container is known to be an exact
%% list (or array or tuple IIRC): when we see e.g. `for item in mylist` and we
%% know statically that `mylist` is an exact list/tuple/array, instead of
%% generating the usual bytecode that goes through the dynamic iterator protocol,
%% we will generate more efficient bytecode that uses `FAST_LEN` to get the length
%% and then uses an incremented (primitive) index counter and our optimized
%% `SEQUENCE_GET` opcode (without bounds checking since in this case we can
%% guarantee the index counter can never be out of bounds). The visible semantics
%% remain the same but under the hood we use numeric indexing instead of the
%% iterator protocol. So this optimization is much more generally applicable, and
%% relies on FAST_LEN.
%%
%% Also I think FAST_LEN is a convenient example optimization for explication
%% purposes, because it is simple and easy to understand but it’s reasonably
%% representative of the general approach.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
module imports, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's random testing tools~\cite{kf-sfp-2009} to check
type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
%% FILL how many programs exactly?
%% FILL what bugs did testing catch?
Finally, we employed the well-established method of testing end-to-end
conformance~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}
with a test suite.
In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,\footnote{For most of the \numSPtest{} tests, the translation is automatic.
  A few tests required hand-pruning to remove features that the model
  does not handle (\totalnum{52}).
  \SP{} has \numSPdiff{} other tests (\numSPtotal{} total) that we did not use
  because they fall outside the scope of the model.
  \Cref{s:impl} summarizes the out-of-scope features;
  \cref{a:banned-test} gives a detailed categorization.
  }
which suggests that the model conforms to actual \SP{}.

@@

\section{Skipped \SP{} Regression Tests}
\label{a:banned-test}

As of January 9 (commit \href{https://github.com/facebookincubator/cinder/commit/6d61575e28b2d1f31a197f883486c5f012de98c1}{6d61575}),
the \SP{} test suite contains \numSPtotal{} tests.
We use \numSPtest{} of these tests for the model.
The remaining \numSPdiff{} tests are skipped for the following reasons:

\begin{itemize}
  \item TBD need categories
\end{itemize}

\section{Fine-Grained Benchmark Data}
\label{a:microbench-detail}

@@

\begin{center}
\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}
\end{center}
